---
title: "Artemisia annua RNA-seq Project: Differential Expression Analysis"
author: "Nyssa Ndey-Bongo"
date: today
format: 
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: false
    code-fold: true
    code-tools: true
    theme: flatly
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This notebook performs differential expression analysis on the _Artemisia annua_ samples taken from the Zhang et al. 2018. It does so using DESeq2. 

Analysis code was borrowed from KAUST Academy. 

Source: [https://bioinfo-kaust.github.io/academy-stage3-2026/html/lab5.html](https://bioinfo-kaust.github.io/academy-stage3-2026/html/lab5.html)

**INPUTS:**
- counts/tximport_object.rds       — the tximport object with gene counts
- samples.csv                      — sample metadata with conditions

**OUTPUTS:**
- deseq2_all_results.csv      — table of all DESeq2 results (log2FC, p-value)
- deseq2_results.rds          — the full DESeq2 object for downstream analysis
- deseq2_significant.csv      — table of all signficant DESeq2 results (log2FC, p-value)
- normalized_counts.csv       — normalized gene count matrix
- deseq2_summary.txt          — summary of DESeq2 results (number of DE genes, top hits, etc.)

## Setup
```{bash}
conda activate genomics
```

## Load Packages
```{r}
#| label: load-packages

library(DESeq2)      # performs differential gene expression analysis
library(tximport)    # loads and reads tximport object
library(tidyverse)   # string manipulation

```


## Load Data
```{r}
#| label: load-data

txi <- readRDS("../counts/tximport_object.rds")
sample_info <- read.csv("../samples.csv", header = TRUE, sep = ",")

# Set factor levels
sample_info$condition <- factor(
    sample_info$condition,
    levels=c("white", "red", "far_red", "blue")
)

cat(paste0("Samples: ", nrow(sample_info), "\n"))
cat(paste0("Genes: ", nrow(txi$counts), "\n"))
cat(paste0("Conditions: ", paste(levels(sample_info$condition), collapse = " vs "), "\n"))
```


## Pre-Filter
```{r}
#| label: pre-filter

# creates a DESeq2 dataset object from tximport data
dds <- DESeqDataSetFromTximport(
    txi, 
    colData = sample_info,   # samples.csv becomes the metadata
    design = ~condition      # 'condition' is a column in samples.csv
)

# NOTE: condition is the grouping variable DESeq2 uses for differential
# expression models. replicate is a useful metadata to have. r1 and r2
# are ignored by DESeq2, it only cares about non-path columns
# Batch correction or additional covariates like ~ batch + condition,
# should be added as columns to samples.csv before running 
# DESeq2 analysis. 

# Pre-filtering: Remove genes with very low counts
keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]
cat(paste0("Genes after filtering (>=10 counts in >=3 samples): ", nrow(dds), "\n"))
```

## Analysis
```{r}
#| label: analysis

# Begin running DESeq on the DESeq dataset object
dds <- DESeq(dds)

# --- Obtain results from desired comparisons ---

# White versus red
res_wl_vs_red     <- results(
    dds, contrast = c("condition", "red", "white"), 
    name = "condition_white_vs_red"
)
res_wl_vs_red     <- res_wl_vs_red[order(res_wl_vs_red$padj), ]

summary(res_wl_vs_red)

# White versus blue
res_wl_vs_blue    <- results(
    dds,
    contrast = c("condition", "blue", "white"),
    name = "condition_white_vs_blue"
)
res_wl_vs_blue    <- res_wl_vs_blue[order(res_wl_vs_blue$padj), ]

summary(res_wl_vs_blue)

# White versus far-red
res_wl_vs_far_red <- results(
    dds,
    contrast = c("condition", "far_red", "white"),
    name = "condition_white_vs_far_red"
)
res_wl_vs_far_red <- res_wl_vs_far_red[order(res_wl_vs_far_red$padj), ]

summary(res_wl_vs_far_red)

# Place results into a list
comparison_list <- list(res_wl_vs_red, res_wl_vs_blue, res_wl_vs_far_red)
```

## Results
```{r}
#| label: results

# Creating empty lists for all results and significant results
comparison_df_all <- list()
comparison_df_sig <- list()

# Looping through each comparison to obtain results
# Places each comparison result as a discrete dataframe
# Dataframes for each comparison stored in a list

# Log and save all results
for (i in comparison_list) {
    result <- as.data.frame(i)
    results$gene_id <- rownames(result)
    result <- result[, c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")]
    append(comparison_df_all, result)
}

# Log and save significance results
for (i in comparison_list) {
    # Significant: padj < 0.05 and |log2FC| > 1
    sig_genes <- subset(i, padj < 0.05 & abs(log2FoldChange) > 1)
    sig_genes <- sig_genes[order(sig_genes$padj), ]

    sig_res <- as.data.frame(sig_genes)
    sig_res$gene_id <- rownames(sig_res)
    sig_res <- sig_res[, c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")]
    append(comparison_df_sig, sig_res)

    cat(paste0("Total significant genes: ", nrow(sig_genes), "\n"))
    cat(paste0("Upregulated in ",  str_extract(i$name, "_*$"), ":", sum(sig_genes$log2FoldChange > 0, na.rm = TRUE), "\n"))
    cat(paste0("Downregulated in ",str_extract(i$name, "_*$"), ":", sum(sig_genes$log2FoldChange < 0, na.rm = TRUE), "\n"))
}

# Looping through list of dataframes and appending them as one
# Write appended dataframe to one single CSV file
for (df in comparison_df_all) {
    all_res <- rbind(all_res, df)
}

for (df in comparison_df_sig) {
    all_sig <- rbind(all_sig, df)
}

write.csv(all_res, "../results/tables/deseq2_all_results.csv", row.names = FALSE)
write.csv(all_sig, "../results/tables/deseq2_significant.csv", row.names = FALSE)

# Normalized counts
norm_counts <- as.data.frame(counts(dds, normalized = TRUE))
norm_counts$gene_id <- rownames(norm_counts)
write.csv(
    norm_counts,
    glue("{opt$output_dir}/tables/normalized_counts.csv"),
    row.names = FALSE
)
log_msg(paste("Normalized counts saved to:", glue("{opt$output_dir}/tables")))

# Save DESeq2 object for later use
saveRDS(dds, glue("{opt$output_dir}/deseq2_object.rds"))
log_msg(paste("dds object saved to:", opt$output_dir))
```


## Summary
```{r}
#| label: summary
n_genes <- nrow(txi$counts)
n_samples <- nrow(sample_info)

summary_text <- paste0(
    "=== DESeq2 import summary ===\n",
    "Samples        : ", n_samples,    "\n",
    "Genes          : ", n_genes,  "\n",
    "Conditions : ",
        paste(levels(sample_info$condition),
        collapse = " vs "), "\n",
    "========================\n"
)

cat(summary_text)

# save the summary to a text file
writeLines(summary_text, "..logs/deseq2_import_summary.txt")
```

## Notes: 

### Why DESeq2?
DESeq2 is the gold-standard tool for differential expression analysis because it:
- Models count data with negative binomial distribution
- Performs size factor normalization to account for library size differences
- Uses shrinkage estimators for dispersion and fold changes
- Properly controls false discovery rate (FDR) with Benjamini-Hochberg correction

### Key Concepts
| Term | Description |
|------|-------------|
| log2 Fold Change | log2(KO/WT) - positive means upregulated in KO |
| p-value | Probability of observing the difference by change |
| padj (adjusted p-value) | p-value corrected for multiple testing (FDR) |
| baseMean | Average normalized count across all samples | 